Distributional Trees for Circular Data

### 2019-02-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
adaptions following formatting guidelines
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2441 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: eb2f6fbdb34d0043eaed171a3ad052491561e4aa

% Keywords (at most 5):
\keywords{Distributional trees; Circular response; Von Mises distribution.}

@@

In this study distributional trees are adapted to circular responses by employing the von Mises
distribution, also known as ``the circular normal distribution''. Based on a location parameter
$\mu \in [-\pi, \pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [-\pi, \pi]$
is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
(see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).

@@

\begin{figure}[p!]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates employed are
wind direction (dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$),
and pressure gradients along the upper (pdiff\_up; $\text{hPa}$) and lower
(pdiff\_low; $\text{hPa}$) Inn Valley.}
\label{fig:tree} \end{figure}

@@

Figure~\ref{fig:tree} depicts the resulting distributional tree, including
both the empirical (gray) and fitted von Mises (red) distribution of wind direction
in each terminal node. Based on the fitted location parameters $\hat \mu$, the subgroups
can be distinguished into the following wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley
(from east to west, nodes 4 and 5). (2) Downslope winds blowing across
the Alpine crest along the Wipp Valley towards Innsbruck (from south-east to
north-west, nodes 7 and 8). (3) Down-valley winds blowing in the direction
of the valley mouth (from west to east, nodes 12, 14, and 15). Node~11
captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.

@@

A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model
(see Schlosser et.~al, 2018 for a general discussion of distributional forests).

Another possible improvement for obtaining more parsimonious and more stable trees is to
consider splitting circular covariates into two circle segments by searching two
split points simultaneously (rather than sequentially at different depths as
for variable ``dd'' in Figure~\ref{fig:tree}).

@@

\item[Jammalamadaka, S. R., and Sengupta, A.] (2001).
     Topics in Circular Statistics.
     {\it World Scientific}.

>>>>

% Keywords (at most 5):
\keywords{Distributional Trees; Circular Response; Von Mises Distribution.}

@@

In this study distributional trees are adapted to circular responses by employing the von Mises
distribution, also known as ``the circular normal distribution''. Based on a location parameter
$\mu \in [-\pi, \pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [-\pi, \pi]$
is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
(see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).

@@

\begin{figure}[p!]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{schlosser-circtree_plot.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates employed are
wind direction (dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$),
and pressure gradients along the upper (pdiff\_up; $\text{hPa}$) and lower
(pdiff\_low; $\text{hPa}$) Inn Valley.}
\label{schlosser:fig_tree} \end{figure}

@@

Figure~\ref{schlosser:fig_tree} depicts the resulting distributional tree, including
both the empirical (gray) and fitted von Mises (red) distribution of wind direction
in each terminal node. Based on the fitted location parameters $\hat \mu$, the subgroups
can be distinguished into the following wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley
(from east to west, nodes 4 and 5). (2) Downslope winds blowing across
the Alpine crest along the Wipp Valley towards Innsbruck (from south-east to
north-west, nodes 7 and 8). (3) Down-valley winds blowing in the direction
of the valley mouth (from west to east, nodes 12, 14, and 15). Node~11
captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.

@@

A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model
(see Schlosser et.~al~(2018) for a general discussion of distributional forests).

Another possible improvement for obtaining more parsimonious and more stable trees is to
consider splitting circular covariates into two circle segments by searching two
split points simultaneously (rather than sequentially at different depths as
for variable ``dd'' in Figure~\ref{schlosser:fig_tree}).

@@

\item[Jammalamadaka, S. R. and Sengupta, A.] (2001).
     {\it Topics in Circular Statistics}.
     World Scientific.
